Eliminating Systematic Bias in Bright-Siren Multimessenger Observations of Compact Object Mergers from Inclination-Dependent Emission

### 2022-10-03 src/tex/ms.tex
Compare structured emission to isotropic (with more sources).

Different EM counterparts will exhibit different dependence on inclination
angle.  In this subsection, we model an GW+EM catalog composed of two
components.  One component has EM detectable emission only very close to on/off
axis, with $P_{\detected,\EM} = 0$ unless $|x| \geq x_\mathrm{min}$ with
$x_\mathrm{min} = \cos 15^\circ$.  The other component has a (weak) preference
for on/off axis emission, with detectability varying quadratically from a
maximum at $|x| = 1$ to and 1/3 of its maximum value at $x = 0$.  The two
components are mixed equally in the data set, so that the overall EM detection
efficiency is 
\begin{equation}
    P_{\detected,\EM}(x) \propto \frac{1}{2} \left( \frac{H\left( |x| - x_\mathrm{min} \right)}{x_\mathrm{min}} + \frac{3 \left( x^2 + 1/2 \right)}{5} \right),
\end{equation}
with $H$ the Heaviside step function.  We choose $N_l = 4$.  The logarithm of
this function is not expressible as any finite sum of Legendre polynomials, so
our model is incapable of exactly reproducing the shape of this detection
efficiency for any choice of $N_l$.  Nonetheless, we will see that we can still
obtain sub-percent accurate, un-biased estimates of $H_0$, even with only four
additional parameters to describe the EM selection function.

@@

\begin{figure}
    \script{H0-top-hat.py}
    \includegraphics[width=\columnwidth]{figures/H0-top-hat.pdf}
    \caption{Posterior on $H_0$ from fitting the mock catalog described in \S\,
    \ref{sec:strongly-variable}.  The vertical black line shows $H_0 = 0.7$, the
    value used to prepare the catalog.  The blue curve shows the posterior when
    fitting $P_{\detected,\EM}$ using Eq.\ \eqref{eq:Pdet-model} with $N_l=4$.
    In this case we find $H_0 = \variable{output/H0-top-hat.txt}$ (median and
    68\% credible interval).  The orange curve shows the posterior when fitting
    without accounting for inclination-dependent EM selection.  In this case we
    find $H_0 = \variable{output/H0-top-hat-flat.txt}$, with $\sim 8 \sigma$
    bias.}
    \label{fig:H0-top-hat}
\end{figure}

>>>>

Different EM counterparts will exhibit different dependence on inclination
angle.  In this subsection, we model an GW+EM catalog composed of two
components.  One component has EM detectable emission only very close to on/off
axis, with $P_{\detected,\EM} = 0$ unless $|x| \geq x_\mathrm{min}$ with
$x_\mathrm{min} = \cos 15^\circ$.  The other component has a (weak) preference
for on/off axis emission, with detectability varying quadratically from a
maximum at $|x| = 1$ to and 1/3 of its maximum value at $x = 0$.  The two
components are mixed equally in the population of binary mergers, so that the
overall EM detection efficiency is 
\begin{equation}
    P_{\detected,\EM}(x) \propto \frac{1}{2} \left( \frac{H\left( |x| - x_\mathrm{min} \right)}{x_\mathrm{min}} + \frac{3 \left( x^2 + 1/2 \right)}{5} \right),
\end{equation}
with $H$ the Heaviside step function.  We choose $N_l = 4$.  The logarithm of
this function is not expressible as any finite sum of Legendre polynomials, so
our model is incapable of exactly reproducing the shape of this detection
efficiency for any choice of $N_l$.  Nonetheless, we will see that we can still
obtain sub-percent accurate, un-biased estimates of $H_0$, even with only four
additional parameters to describe the EM selection function.

@@

\begin{figure}
    \script{H0-top-hat.py}
    \includegraphics[width=\columnwidth]{figures/H0-top-hat.pdf}
    \caption{Posterior on $H_0$ from fitting the mock catalog described in \S\,
    \ref{sec:strongly-variable}.  The vertical black line shows $H_0 = 0.7$, the
    value used to prepare the catalog.  The blue curve shows the posterior when
    fitting $P_{\detected,\EM}$ using Eq.\ \eqref{eq:Pdet-model} with $N_l=4$.
    In this case we find $H_0 = \variable{output/H0-top-hat.txt}$ (median and
    68\% credible interval).  The orange curve shows the posterior when fitting
    without accounting for inclination-dependent EM selection.  In this case we
    find $H_0 = \variable{output/H0-top-hat-flat.txt}$, with $\sim 6 \sigma$
    bias.}
    \label{fig:H0-top-hat}
\end{figure}

With this choice of EM detection efficiency, we observe a counterpart in about
15\% of the GW-detectable systems.  It is interesting to ask whether the
improvement in inclination estimation from the structured EM emission (and
therefore improved distance estimation from the GW observations) outweighs the
loss of sources compared to a scenario where the EM emission is uniform.  (In
other words, would it be better to employ a search strategy that selects sources
that are likely to have anisotropic EM emission in order to improve distance
estimates?)  For this distribution, a catalog with $1/0.15 \sim 7$ times as many
sources with isotropic EM emission provides overall a better Hubble parameter
measurement, as shown in Figure \ref{fig:top-hat-flat-compare}.  (For both mock
catalogs, we employ $N_l = 4$, allowing for the same possibility of structure in
the EM emission; the inferred $P_{\detected,\EM}$ for the catalog with isotropic
emission is shown in Figure \ref{fig:flat-Pdet}.)

\begin{figure}
    \script{top-hat-flat-compare.py}
    \includegraphics[width=\columnwidth]{figures/top-hat-flat-compare.pdf}
    \caption{Comparison between $H_0$ inferred from a mock catalog with the
    anisotropic EM emission discussed in \S\ \ref{sec:strongly-variable} (blue
    curve) with a catalog with an equivalent GW merger rate but isotropic
    emission (and therefore $\sim 7$ times as many detections; orange curve).
    The improved catalog size from the isotropic emission overcomes the improved
    distance estimates from the knowledge of inclination gained from the
    anisotropic emission, and the overall measurement is tighter.  We find the
    posterior median and 68\% credible interval $H_0 =
    \variable{output/H0-top-hat.txt}$ from the catalog with anisotropic EM
    emission, and $H_0 = \variable{output/H0-flat-comparison.txt}$ from the
    larger catalog with isotropic emission.}
    \label{fig:top-hat-flat-compare}
\end{figure}

\begin{figure}
    \script{flat-Pdet.py}
    \includegraphics[width=\columnwidth]{figures/flat-Pdet.pdf}
    \caption{The inferred inclination distribution for the catalog with
    isotropic EM emission and equivalent GW merger to the anisotropic emission
    described in \S\ \ref{sec:strongly-variable}.  The blue line gives the
    posterior median and the dark and light bands the posterior 68\% and 95\%
    credible interval.  Even though the EM emission is isotropic, we fit the
    detection efficiency with $N_l = 4$ for a fair comparison with the (smaller)
    catalog of anisotropic sources whose inferred inclination distribution is
    shown in Figure \ref{fig:top-hat-Pdet}.}
    \label{fig:flat-Pdet}
\end{figure}
